---
title: "01 - Analysis 1"
author: "Selin Jessa [[selin.jessa@mail.mcgill.ca](mailto:selin.jessa@mail.mcgill.ca)]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: flatly
    css: ../include/style.css
    toc: yes
    toc_depth: 4
    number_sections: true
    df_print: paged
    includes:
      before_body: ../include/header.html
      after_body:  ../include/footer.html
---

<!-- FRONT MATTER, insert configuration info -->

```{r header, echo = FALSE, results = 'asis', warning = FALSE}

# Index of the document
# ...determines name of the subfolder of `outputs` and `figures`
doc_id <- "01"

suppressMessages(library(here))

# Knit child document with header
res <- knitr::knit_child(here("include", "header.Rmd"), envir = environment(), quiet = TRUE)
cat(res, sep = '\n')

```

<!-- END OF FRONT MATTER -->


# Overview

This is an example of the first analysis in a project.

# Libraries

```{r libs, warning = FALSE, message = FALSE, echo = TRUE, cache = FALSE}

library(ggplot2)
library(tidyverse)

```

```{asis, echo = TRUE}

Here is a note to self. This markdown code is contained in a chunk, 
with visibility controlled by the `echo` chunk option.

Change echo to TRUE or FALSE to display/hide this chunk.

```


# Analysis

## Part 1

This part of the analysis generates a figure; we can use the chunk option
`echo_fig = TRUE` to print its output path.

```{r pressure, echo_fig = TRUE}

mtcars %>% 
    ggplot(aes(x = mpg, y = cyl)) +
    geom_point() +
    theme_bw()

```

## Part 2

This part of the analysis saves the `mtcars` dataset from R as a TSV.
By matching the chunk name to the output file name, we can then use
the `echo_out = TRUE` chunk option to show the path to the file.

```{r mtcars.tsv, echo_out = TRUE}

head(mtcars)

rr_write_tsv(df = mtcars,
             path = file.path(out, "mtcars.tsv"),
             desc = "The mtcars dataset, verbatim")

```


<!-- END MATTER, insert reproducibility info -->

```{r footer, echo = FALSE, results = 'asis', warning = FALSE, cache = FALSE}

# Knit child document with header
res <- knitr::knit_child(here("include", "footer.Rmd"), envir = environment(), quiet = TRUE)
cat(res, sep = '\n')

```

<!-- END OF END MATTER -->
